---
title: "Buoy_Students"
format: pdf
editor: visual
---

## Packages to Install

```{r, warning = FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(zoo)
library(tibble)
library(readr)
```

## Compiling all years' data

1.  Try to remember why 2007 is the split year
2.  Why are using different functions to read data prior and post 2007?
3.  What are some new function you got to know from following code? what do these functions do?

```{r, warning=FALSE, message=FALSE}
file_root <- "https://www.ndbc.noaa.gov/view_text_file.php?filename=44013h"
tail <- ".txt.gz&dir=data/historical/stdmet/"

load_buoy_data <- function(year) {
  path <- paste0(file_root, year, tail)

  header <- scan(path, what = 'character', nlines = 1)
  num_columns <- length(header)  

  if (num_columns == 16) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, mm = NA, .after = "hh")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else if (num_columns == 17) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else {
    buoy <- fread(path, header = FALSE, skip = 1, fill = TRUE)
    setnames(buoy, header)
  }

  return(buoy)
}
all_data <- lapply(1985:2024, load_buoy_data)
combined_data <- rbindlist(all_data, fill = TRUE)
```

## Cleaning and Organizing the data

We start by merging all different version of `Year` column. We do same with other columns which are same but having data for certain set of years. We remove the remaining columns after merging them.

Creating datetime column using `lubridate()`

```{r}
combined_data <- combined_data %>%
  mutate(
    YY = as.character(YY),
    `#YY` = as.character(`#YY`),
    YYYY = as.character(YYYY)
  )

# Combine year columns safely using coalesce
combined_data <- combined_data %>%
  mutate(YYYY = coalesce(YYYY, `#YY`, YY))
combined_data <- combined_data %>%
  mutate(BAR = coalesce(as.numeric(BAR), as.numeric(PRES)),  # Convert BAR and PRES to numeric
    WD = coalesce(as.numeric(WD), as.numeric(WDIR)))

combined_data <- combined_data %>%
  select(-TIDE, -TIDE.1, -mm,- WDIR, -PRES,-`#YY`,-YY)

combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))

combined_data <- combined_data %>%
  mutate(across(everything(), 
                ~ na_if(as.numeric(as.character(.)), 99) %>%
                na_if(999) %>%
                na_if(9999)))

summary(combined_data)
#str(combined_data)
#str(combined_data$datetime)
if (!inherits(combined_data$datetime, "POSIXct")) {
  combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))
}
```

```{r}
combined_data <- combined_data %>%
  mutate(Year = year(datetime))

combined_data <- combined_data %>% select(-YYYY)

yearly_avg_temp <- combined_data %>%
  group_by(Year) %>%
  summarise(
    avg_air_temp = mean(ATMP, na.rm = TRUE),
    avg_water_temp = mean(WTMP, na.rm = TRUE)
  )

# Ensure Year is numeric
yearly_avg_temp$Year <- as.numeric(as.character(yearly_avg_temp$Year))

ggplot(yearly_avg_temp, aes(x = Year)) +
  geom_line(aes(y = avg_air_temp, color = "Air Temperature"), size = 1) +
  geom_line(aes(y = avg_water_temp, color = "Water Temperature"), size = 1) +
  scale_x_continuous(
    breaks = seq(min(yearly_avg_temp$Year, na.rm = TRUE),
                 max(yearly_avg_temp$Year, na.rm = TRUE),
                 2)   # every year
  ) +
  labs(
    title = "Temperature Trends Over Time",
    x = "Year",
    y = "Temperature (°C)",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

#Under this line is my assignment D work

Question: Is there a correlation between WSPD and WVHT?
```{r}
yearly_avg_data <- combined_data %>%
  group_by(Year) %>%
  summarise(
    avg_WSPD = mean(WSPD, na.rm = TRUE),
    avg_WVHT = mean(WVHT, na.rm = TRUE),
    avg_BAR = mean(BAR,na.rm = TRUE)
    )

ggplot(data = yearly_avg_data,aes(x = Year))+
  geom_line(aes(y = avg_WSPD,color = "WSPD"),size = 1) +
  geom_line(aes(y = avg_WVHT,color = "WVHT"),size = 1)
```
We can see a roughly positive relationship between these two elements.


To examine this more closely, we focused on 2024 data.
We need to use scatter point and linear model to find out the result based on 2024 data
```{r}
df2024 <- read.csv("2024.txt",sep = "",header = TRUE)
df2024 <- df2024[-1,]
df2024 <- df2024 %>%
  mutate(
    date_label = paste0(as.integer(MM), ".", as.integer(DD)),
    WSPD = as.numeric(WSPD),
    WVHT = as.numeric(WVHT)
  )
#clean up the data set and make wspd and wvht numeric 

df2024_daily <- df2024 %>%
  group_by(date_label) %>%
  summarise(
    avg_WSPD = mean(WSPD, na.rm = TRUE),
    avg_WVHT = mean(WVHT, na.rm = TRUE),
    )

fit <- lm(avg_WVHT ~ avg_WSPD,data = df2024_daily)
summary(fit)


ggplot(df2024_daily, aes(x = avg_WSPD, y = avg_WVHT)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "WVHT vs WSPD",x = "Wind speed",y = "Wave height")
```
From the plot and regression model, we can tell there's a positive relationship between WVHT and WSPD. 
However, wind speed alone explains only about 17% of the variance, 
so other factors beyond only wind speed are important in determining observed wave heights.


