---
title: "Tingji-AssignmentD"
format: html
editor: visual
---

```{r, warning = FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(zoo)
library(tibble)
library(readr)
```

These are the given code from Buoy_students.qmd
```{r, warning=FALSE, message=FALSE}
file_root <- "https://www.ndbc.noaa.gov/view_text_file.php?filename=44013h"
tail <- ".txt.gz&dir=data/historical/stdmet/"

load_buoy_data <- function(year) {
  path <- paste0(file_root, year, tail)

  header <- scan(path, what = 'character', nlines = 1)
  num_columns <- length(header)  

  if (num_columns == 16) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, mm = NA, .after = "hh")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else if (num_columns == 17) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else {
    buoy <- fread(path, header = FALSE, skip = 1, fill = TRUE)
    setnames(buoy, header)
  }

  return(buoy)
}
all_data <- lapply(1985:2024, load_buoy_data)
combined_data <- rbindlist(all_data, fill = TRUE)
```

```{r}
#| messages: false
#| warning: false
combined_data <- combined_data %>%
  mutate(
    YY = as.character(YY),
    `#YY` = as.character(`#YY`),
    YYYY = as.character(YYYY)
  )

# Combine year columns safely using coalesce
combined_data <- combined_data %>%
  mutate(YYYY = coalesce(YYYY, `#YY`, YY))
combined_data <- combined_data %>%
  mutate(BAR = coalesce(as.numeric(BAR), as.numeric(PRES)),  # Convert BAR and PRES to numeric
    WD = coalesce(as.numeric(WD), as.numeric(WDIR)))

combined_data <- combined_data %>%
  select(-TIDE, -TIDE.1, -mm,- WDIR, -PRES,-`#YY`,-YY)

combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))

combined_data <- combined_data %>%
  mutate(across(everything(), 
                ~ na_if(as.numeric(as.character(.)), 99) %>%
                na_if(999) %>%
                na_if(9999)))

if (!inherits(combined_data$datetime, "POSIXct")) {
  combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))
}
combined_data <- combined_data %>%
  mutate(Year = year(datetime))

combined_data <- combined_data %>% select(-YYYY)
```
<p style="page-break-after:always">

#Start of my Assignment D
Question: Is there a correlation between WSPD and WVHT?
```{r}
#| messages: false
#| warning: false
yearly_avg_data <- combined_data %>%
  group_by(Year) %>%
  summarise(
    avg_WSPD = mean(WSPD, na.rm = TRUE),
    avg_WVHT = mean(WVHT, na.rm = TRUE),
    avg_BAR = mean(BAR,na.rm = TRUE)
    )

ggplot(data = yearly_avg_data,aes(x = Year))+
  geom_line(aes(y = avg_WSPD,color = "WSPD"),size = 1) +
  geom_line(aes(y = avg_WVHT,color = "WVHT"),size = 1)
```
We can see a roughly positive relationship between these two elements.


To examine this more closely, we focused on 2024 data.
We need to use scatter point and linear model to find out the result based on 2024 data
```{r}
df2024 <- read.csv("2024.txt",sep = "",header = TRUE)
df2024 <- df2024[-1,]
df2024 <- df2024 %>%
  mutate(
    date_label = paste0(as.integer(MM), ".", as.integer(DD)),
    WSPD = as.numeric(WSPD),
    WVHT = as.numeric(WVHT)
  )
#clean up the data set and make wspd and wvht numeric 

df2024_daily <- df2024 %>%
  group_by(date_label) %>%
  summarise(
    avg_WSPD = mean(WSPD, na.rm = TRUE),
    avg_WVHT = mean(WVHT, na.rm = TRUE),
    )

fit <- lm(avg_WVHT ~ avg_WSPD,data = df2024_daily)
summary(fit)


ggplot(df2024_daily, aes(x = avg_WSPD, y = avg_WVHT)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "WVHT vs WSPD",x = "Wind speed",y = "Wave height")
```
From the plot and regression model, we can tell there's a positive relationship between WVHT and WSPD. 
However, wind speed alone explains only about 17% of the variance, 
so other factors beyond only wind speed are important in determining observed wave heights.

