---
title: "Buoy_Students"
format: pdf
editor: visual
---

## Packages to Install

```{r, warning = FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(zoo)
library(tibble)
library(readr)
```

## Compiling all years' data

1.  Try to remember why 2007 is the split year
2.  Why are using different functions to read data prior and post 2007?
3.  What are some new function you got to know from following code? what do these functions do?

```{r, warning=FALSE, message=FALSE}
file_root <- "https://www.ndbc.noaa.gov/view_text_file.php?filename=44013h"
tail <- ".txt.gz&dir=data/historical/stdmet/"

load_buoy_data <- function(year) {
  path <- paste0(file_root, year, tail)

  header <- scan(path, what = 'character', nlines = 1)
  num_columns <- length(header)  

  if (num_columns == 16) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, mm = NA, .after = "hh")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else if (num_columns == 17) {
    buoy <- read.table(path, fill = TRUE, header = TRUE, sep = "")
    buoy <- add_column(buoy, TIDE = NA, .after = "VIS")

  } else {
    buoy <- fread(path, header = FALSE, skip = 1, fill = TRUE)
    setnames(buoy, header)
  }

  return(buoy)
}
all_data <- lapply(1985:2024, load_buoy_data)
combined_data <- rbindlist(all_data, fill = TRUE)
```

## Cleaning and Organizing the data

We start by merging all different version of `Year` column. We do same with other columns which are same but having data for certain set of years. We remove the remaining columns after merging them.

Creating datetime column using `lubridate()`

```{r}
combined_data <- combined_data %>%
  mutate(
    YY = as.character(YY),
    `#YY` = as.character(`#YY`),
    YYYY = as.character(YYYY)
  )

# Combine year columns safely using coalesce
combined_data <- combined_data %>%
  mutate(YYYY = coalesce(YYYY, `#YY`, YY))
combined_data <- combined_data %>%
  mutate(BAR = coalesce(as.numeric(BAR), as.numeric(PRES)),  # Convert BAR and PRES to numeric
    WD = coalesce(as.numeric(WD), as.numeric(WDIR)))

combined_data <- combined_data %>%
  select(-TIDE, -TIDE.1, -mm,- WDIR, -PRES,-`#YY`,-YY)

combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))

combined_data <- combined_data %>%
  mutate(across(everything(), 
                ~ na_if(as.numeric(as.character(.)), 99) %>%
                na_if(999) %>%
                na_if(9999)))

summary(combined_data)
#str(combined_data)
#str(combined_data$datetime)
if (!inherits(combined_data$datetime, "POSIXct")) {
  combined_data$datetime <- ymd_h(paste(combined_data$YYYY, combined_data$MM, combined_data$DD, combined_data$hh, sep = "-"))
}
```

```{r}
combined_data <- combined_data %>%
  mutate(Year = year(datetime))

combined_data <- combined_data %>% select(-YYYY)

yearly_avg_temp <- combined_data %>%
  group_by(Year) %>%
  summarise(
    avg_air_temp = mean(ATMP, na.rm = TRUE),
    avg_water_temp = mean(WTMP, na.rm = TRUE)
  )

# Ensure Year is numeric
yearly_avg_temp$Year <- as.numeric(as.character(yearly_avg_temp$Year))

ggplot(yearly_avg_temp, aes(x = Year)) +
  geom_line(aes(y = avg_air_temp, color = "Air Temperature"), size = 1) +
  geom_line(aes(y = avg_water_temp, color = "Water Temperature"), size = 1) +
  scale_x_continuous(
    breaks = seq(min(yearly_avg_temp$Year, na.rm = TRUE),
                 max(yearly_avg_temp$Year, na.rm = TRUE),
                 2)   # every year
  ) +
  labs(
    title = "Temperature Trends Over Time",
    x = "Year", 
    y = "Temperature (°C)", 
    color = "Legend"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

# 
```{r}
yearly_avg_temp2 <- combined_data %>%
  group_by(Year) %>%
  summarise(
    avg_DPD = mean(DPD, na.rm = TRUE),
    avg_APD = mean(APD, na.rm = TRUE),
    avg_MWD = mean(MWD, na.rm = TRUE)

    )
ggplot(data = yearly_avg_temp2,aes(x = Year))+
  geom_line(aes(y = avg_APD,color = "APD"),size = 1) +
  geom_line(aes(y = avg_DPD,color = "DPD"),size = 1) +
  geom_line(aes(y = avg_MWD,color = "MWD"),size = 1) +
  labs(title = "Test title",x = "Years", y = "test Ys")+
  theme_classic()

  

```

